---
title: "Tracking the Novel Coronavirus"
author: "Kevin Lanning, Wilkes Honors College"
date: "Updated `r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

### Overview

The Novel Coronavirus is an emerging health crisis, particularly in Wuhan (a Chinese city larger than New York City) and the surrounding province of Hubei. Keep in mind that in the United States and elsewhere, [most cases are mild](https://www.washingtonpost.com/outlook/2020/02/28/i-have-coronavirus-so-far-it-isnt-that-bad/). There is in short no need to freak out - but do wash your hands frequently, avoid touching your mouths, eyes, and nose, and remember that the inside of your elbow is a great place to deposit your coughs and sneezes.

This code creates simple visualizations using R. The code written to be readily understandable for introductory data science students. You can download it [here](https://github.com/kevinlanning/2019-nCoV/blob/master/novelCorona.Rmd).

The data come from the Johns Hopkins Center for Systems Science and Engineering (JHU/CSSE). 
They have made their data freely available, and as a consequence this script does not use an API or require authorization from Github. Hopkins also provides an up-to-date ArcGIS mapping of their data [here](https://gisanddata.maps.arcgis.com/apps/opsdashboard/index.html#/bda7594740fd40299423467b48e9ecf6). 

```{r setup, echo=FALSE, message = FALSE}
knitr::opts_chunk$set(echo=FALSE, message = FALSE)
library(tidyverse)
library(httr)
library(purrr)
library(magrittr)
library(lubridate)
# for interactive plots
library(plotly)
library(htmlwidgets)
# for pseudo-log transform in plot (allows 0 values)
library(ggallin)
# get list of files
filelist <- GET("https://api.github.com/repos/CSSEGISandData/2019-nCoV/git/trees/master?recursive=1") %>% 
  content() %>% 
# there is probably a more efficient way to reduce this
# list to a set of filenames
  flatten() %>% 
  map ("path") %>%
  flatten() %>%
  tibble() %>% 
  rename(filename = 1) %>% 
  filter(str_detect(filename,".csv") &
           str_detect(filename,"daily")) 
nsheets <- nrow(filelist)
rawGitFiles <- "https://raw.githubusercontent.com/CSSEGISandData/2019-nCoV/master/"
```

```{r readdata, message = FALSE}

# Reading the data

#The Novel Coronavirus data consists of a series of csv files in a Github repository. This combines them into a single sheet in R. 

# variables to retain or create
numvars <- c("Confirmed", "Deaths", "Recovered")
varlist <- c("Province/State", "Country/Region",
             "Last Update", numvars) 
# one cool trick to initialize a tibble
coronaData <- varlist %>%
     map_dfr( ~tibble(!!.x := logical() ) )
# add data from files to tibble
for (i in 1:nsheets) {
  j <- read_csv(paste0(rawGitFiles,filelist$filename[i]))
# if a variable doesn't exist in sheet, add it
  j[setdiff(varlist,names(j))] <- NA
# datetime is formatted inconsistently
# across files, this must be done before merging  
  j %<>% mutate(`Last Update` = 
           parse_date_time(`Last Update`,
                           c('mdy hp','mdy HM',
                             'mdy HMS','ymd HMS'))) %>% 
  select(varlist) 
  coronaData <- rbind(coronaData, j) 
}
```

```{r cleaning}
### Cleaning (wrangling, munging) the data

# Cleaning the data includes not just finding "errors," but adapting it for our own use. It's generally time consuming, as was the case here. The following letters refer to sections of the code below.

# * a - fix a few missing values outside of China for province and country
# * b - the earliest cases, all in China, did not include country
# * c - because province/state is included inconsistently, an unambiguous place variable is created
# * d - reportdate is standardized (above) and renamed
# * e - in some cases, multiple reports are issued for each day. only the last of these is used for each place.
# * f - for dates where no data was supplied, the most recent (previous) data are used
# * g - values of NA for Deaths, Confirmed, and Recovered cases are replaced by zero.
# * h - Prior to Feb 1, reporting for US included only state, since then, city and state. This drops the (duplicated) province/state-only values beginning Feb 1. 

coronaData %<>% 
# a 
  mutate (`Province/State` = case_when(
    (is.na(`Province/State`) & 
       (`Country/Region` == "Australia")) ~ "New South Wales",
    (is.na(`Province/State`) & 
       (`Country/Region` == "Germany")) ~ "Bavaria", 
    TRUE ~ `Province/State`)) %>% 
  mutate (`Country/Region` = case_when(
    `Province/State` == "Hong Kong" ~ "Hong Kong",
    `Province/State` == "Taiwan" ~ "Taiwan",
    `Province/State` == "Washington" ~ "US",
# b
    is.na (`Country/Region`) ~ "Mainland China",
    TRUE ~ `Country/Region`)) %>% 
# c 
  mutate(place = ifelse(is.na(`Province/State`),
                        `Country/Region`,
                        paste0(`Province/State`,", ",
                               `Country/Region`))) %>% 
  mutate(reportDate = 
           date(`Last Update`)) %>% 
  group_by(place,reportDate) %>% 
# e
  slice(which.max(`Last Update`)) %>% 
  ungroup() %>%
# fill in missing dates for each place for time series
# f
  group_by(place) %>% 
  complete(reportDate = seq.Date(min(reportDate),
                                 today(),
                                 by="day")) %>% 
  fill(c(Confirmed,Deaths,Recovered,
         `Country/Region`,`Province/State`)) %>% 
# g
  ungroup() %>% 
  mutate_if(is.numeric, ~replace_na(., 0)) %>% 
# h
  mutate(dropcase = ((!str_detect(`Province/State`,",")) & 
                       (reportDate  > "2020-01-31") &
                       (`Country/Region` == "Canada" | `Country/Region` == "US"))) %>% 
# dplyr called explicitly here because plotly has taken over 'filter'
  dplyr::filter (!dropcase) %>% 
  select(-c(`Last Update`,`Province/State`,`Country/Region`,dropcase)) 

```


```{r threelocations}
### Simplifying the data: China and the rest of the world

# This separates data into three locations, breaking down China into Hubei (Wuhan) and other, then summarizes results:

  coronaDataSimple <- coronaData %>% 
  mutate(country = case_when(
    str_detect(place,"China") ~ "China",
    TRUE ~ "Other countries")) %>% 
  mutate(location = case_when(
    place == "Hubei, Mainland China" ~ "Hubei (Wuhan)",
    country == "China" ~ "Other China",
# what happens when this line is not commented out?
# why is it written this way?
# str_detect(place, "ruise") ~ "Cruise Ship", 
    TRUE ~ "Elsewhere")) %>% 
  group_by(location,reportDate) %>% 
  summarize(Confirmed = sum(Confirmed),
            Deaths = sum(Deaths),
            Recovered = sum(Recovered)) %>% 
  ungroup()
```

### An initial plot

The first plot is simple, including data for only deaths. A caption is added to show the source of the data. The plot includes both the raw data and a smoothed (LOESS) curve.

```{r simpleplot, message= FALSE}
myCaption <- " Data courtesy JHU/CSSE http://bit.ly/ncvdata"
coronaPlot0 <- coronaDataSimple %>% 
  ggplot(aes(x=reportDate, y = Deaths, color = location)) +
  geom_point() + 
  geom_smooth(se = FALSE) +
  labs(caption = myCaption)
coronaPlot0
```

### Adding recovered cases and making the graph interactive

Here, recovered cases and deaths are included (as these are roughly on the same scale). Additional changes are self-evident.

```{r deathsrecovered, message= FALSE}
mySubtitle <- paste0(
         "Recovered cases (solid line) & deaths (dotted) by region.")
myCaption <- " Data courtesy JHU/CSSE http://bit.ly/ncvdata"
coronaPlot1 <- coronaDataSimple %>% 
  ggplot(aes(x=reportDate, color = location)) +
  geom_point(aes(y=Recovered)) + 
  geom_smooth(aes(y=Recovered), se = FALSE,
              linetype = "solid") + 
  geom_point(aes(y=Deaths)) + 
  geom_smooth(aes(y=Deaths), se = FALSE,
              linetype = "dotted") + 
#  scale_y_continuous(trans=pseudolog10_trans) +
#  geom_line(aes(y=Deaths, 
#                color = location), 
#            linetype = "dotted") +
  theme(axis.title.y = 
        element_text(angle = 90,
                     vjust = 1,size = 14),
        legend.position = (c(.2,.8))) +
  labs(title = "Novel coronavirus",
       subtitle = mySubtitle,
       y = "Cases", 
       caption = myCaption)

#Making the graph interactive

# Plotly is an open-source, javascript based library that produces interactive graphs. The syntax that Plotly requires is (a little) different from ggplot, so, for example, the subtitle and caption are folded in to the title here, and the legend is moved a little further over. 

p <- ggplotly(coronaPlot1) %>% 
  # make interactive
  layout(legend = list(x = .1,y = .9),
         title = list(text = paste0('Novel coronavirus',
                                    '<br>',
                                    '<sup>',
                                    mySubtitle,
                                    myCaption,
                                    '</sup>')))
# saveWidget(p, file="coronaDeathsRecovered.html")
p
```

### Plotting confirmed cases

```{r}

#In this figure, data for confirmed cases are shown (only the interactive version is included here). : 
mySubtitle <- paste0(
         "Confirmed cases by region through ",
         (month(today())), "/",         (day(today())),"/",
         (year(today())),".")
coronaPlot2 <- coronaDataSimple %>% 
  ggplot(aes(x=reportDate, color = location)) +
  geom_point(aes(y=Confirmed)) + 
  geom_smooth(aes(y=Confirmed), se = FALSE,
              linetype = "solid") + 
# adding a text block
# The code for this is clumsy
# the cleaner geom_label does not work correctly with plotly
  annotate("rect",ymin=46000,ymax=50000, 
           xmin=date("2020-02-06"), xmax=date("2020-02-12"), 
           fill="white") +
  annotate("text", x=date("2020-02-09"), y=49200, 
           label = "From 2/14, broader",
           size = 3) +
  annotate("text", x=date("2020-02-09"), y=46800, 
           label = "diag. criteria are used",
           size = 3) +
  theme(axis.title.y = 
        element_text(angle = 90,
                     vjust = 1,size = 14),
        legend.position = (c(.2,.8))) +
  labs(title = "Novel coronavirus",
       subtitle = mySubtitle,
       y = "Cases", 
       caption = myCaption)

p <- ggplotly(coronaPlot2) %>% 
  # make interactive
  layout(legend = list(x=.1,y=.9),
         title = list(text = paste0('Novel coronavirus',
                                    '<br>',
                                    '<sup>',
                                    mySubtitle,
                                    myCaption,
                                    '</sup>')))
p



```


### Looking at the rate of growth in Wuhan

To examine whether the growth rate is exponential, data for all three measures (Confirmed, Deaths, and Recovered) are plotted for Wuhan using a log scale.  Exponential growth is indicated by a linear trend. 

```{r logplots, message= FALSE}

mySubtitle <- paste0(
         "Recovered cases (solid line), deaths (dotted), and confirmed (dashed) in Wuhan through ",
         (month(today())), "/",
         (day(today())),"/",
         (year(today())),".")
myCaption <- " Data courtesy JHU/CSSE http://bit.ly/ncvdata"
coronaPlot3 <- coronaDataSimple %>% 
  dplyr::filter(location == "Hubei (Wuhan)") %>% 
  ggplot(aes(x=reportDate)) +
# three variables
  geom_point(aes(y=Recovered, color = "Recovered")) + 
  geom_smooth(aes(y=Recovered, color = "Recovered"),
              se = FALSE, linetype = "solid") + 
  geom_point(aes(y=Deaths, color = "Deaths")) + 
  geom_smooth(aes(y=Deaths, color = "Deaths"),
              se = FALSE, linetype = "dotted") + 
  geom_point(aes(y=Confirmed, color = "Confirmed")) + 
  geom_smooth(aes(y=Confirmed, color = "Confirmed"),
              se = FALSE, linetype = "dashed") + 
# log scale (pseudo becaue log 0 is inf)
  scale_y_continuous(trans=pseudolog10_trans) +
  theme(axis.title.y = 
        element_text(angle = 90,
                     vjust = 1,size = 14),
        legend.position = (c(.8,.2))) +
# kludge to change legend title
    scale_colour_discrete("Case Type") +
  labs(title = "Novel coronavirus",
       fill = "Case type",
       subtitle = mySubtitle,
       y = "Cases", 
       caption = myCaption)
ggplotly(coronaPlot3) %>% 
  layout(legend = list(x=.8,y=.2),
         title = list(text = paste0('Novel coronavirus',
                                    '<br>',
                                    '<sup>',
                                    mySubtitle,
                                    myCaption,
                                    '</sup>')))
```

### Looking at growth rates *outside of* China.

In the last plot, data for all three measures (Confirmed, Deaths, and Recovered) are plotted for cases outside of China using a log scale. 

```{r logplotsoutsideChina, message= FALSE}

mySubtitle <- paste0(
         "Recovered cases (solid line), deaths (dotted), and confirmed (dashed) outside of China through ",
         (month(today())), "/",
         (day(today())),"/",
         (year(today())),".")
myCaption <- " Data courtesy JHU/CSSE http://bit.ly/ncvdata"
coronaPlot3 <- coronaDataSimple %>% ungroup () %>%  
  dplyr::filter(location == "Elsewhere") %>% 
  ggplot(aes(x=reportDate)) +
# three variables
  geom_point(aes(y=Recovered, color = "Recovered")) + 
  geom_smooth(aes(y=Recovered, color = "Recovered"),
              se = FALSE, linetype = "solid") + 
  geom_point(aes(y=Deaths, color = "Deaths")) + 
  geom_smooth(aes(y=Deaths, color = "Deaths"),
              se = FALSE, linetype = "dotted") + 
  geom_point(aes(y=Confirmed, color = "Confirmed")) + 
  geom_smooth(aes(y=Confirmed, color = "Confirmed"),
              se = FALSE, linetype = "dashed") + 
# log scale (pseudo becaue log 0 is inf)
  scale_y_continuous(trans=pseudolog10_trans) +
  theme(axis.title.y = 
        element_text(angle = 90,
                     vjust = 1,size = 14),
        legend.position = (c(.8,.2))) +
# kludge to change legend title
    scale_colour_discrete("Case Type") +
  labs(title = "Novel coronavirus",
       fill = "Case type",
       subtitle = mySubtitle,
       y = "Cases", 
       caption = myCaption)
ggplotly(coronaPlot3) %>% 
  layout(legend = list(x=.8,y=.2),
         title = list(text = paste0('Novel coronavirus',
                                    '<br>',
                                    '<sup>',
                                    mySubtitle,
                                    myCaption,
                                    '</sup>')))
```

### Some questions

1. Grab the code and try to run it yourself. What problems (if any) did you encounter? 

2. Can you reverse-engineer the code? Where is it confusing? (remember the 15 minute rule).

3. What is the relationship between 'confirmed cases' and 'deaths'? Which appears to be increasing more quickly?  Is this cause for optimism?

4. Can you improve on these plots? 

### Additional notes

If you are interested in looking at additional epidemiological datasets and how they might be looked at in R, consider this source by Tomás J. Aragón (https://bookdown.org/medepi/phds/). For Plotly in R, check out https://plotly-r.com/ 

This document was modified February 3 because of new Googlesheet link and altered variable names, on Feb 5 because of a new URL for the data and additional changes in the variable name for date, and Feb 7 to (a) remove need for OAuth and (b) separate Wuhan from other China. On Feb 9, additional data cleaning was performed and interactive plots were added. On February 11, the code was rewritten to read files from a Github repo rather than Google Sheets. Minor changes and additions were made in late February and early March.